---
title: "Defining the model"
author: Marcello Gallucci
date: '`r Sys.Date()`'
output: 
     rmarkdown::html_vignette:
          css: style.css
     html_document:
         toc: true
         toc_float:
             collapsed: false
         css: style.css
vignette: >
  %\VignetteIndexEntry{Defining the model}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "",
  fig.align = "center",
  fig.height = 4,
  fig.width = 6
)
options(digits=3)

```

```{r setup}
library(GAMLj3)
data("clustermanymodels")
data<-clustermanymodels
```

# Main model

## Model formula

The easiest and recommended way to define a model is to pass a formula. Formula formats accepted by `stats:glm` and `lme4::lmer()` functions are accepted.

For instance, a regression with interaction is passed as:

```{r warning=FALSE, message=FALSE, results='hide', eval=FALSE}
mod1<-gamlj_lm(formula = y~x*z, data=data)
```

For mixed and generalized mixed models the `lme4::lmer()` format is accepted:

```{r warning=FALSE, message=FALSE, results='hide', eval=FALSE}
mod2<-gamlj_lm(formula = y~x*z+(1+x|cluster), data=data)
```

For logistic models, a part from the standard format, one can pass the formula with two additional format: The first allows passing the counts of successes and the counts of failures as a combined array.

```{r warning=FALSE, message=FALSE, results='hide', eval=FALSE}
mod3<-gamlj_glm(formula = cbind(success,failure)~x*z, model_type="logistic",data=data)
```

Alternatively, one can pass the probability of success (with $0\le p \le 1$) and the counts of cases ($tot$) for each row in the data.frame. The format uses `/` to indicate `p over T`.

```{r warning=FALSE, message=FALSE, results='hide', eval=FALSE}
mod4<-gamlj_glm(formula = p / tot ~x*z, model_type="logistic", data=data)
```

In all models, polynomial terms can be passed using the `I()` format:

```{r warning=FALSE, message=FALSE, results='hide', eval=FALSE}
mod5<-gamlj_lm(formula = y ~x + I(x^2) , data=data)
```

Please notice that the `I()` is not interpreted literally, so the term is build after any manipulation of the variable takes places, such as centering or standardizing.   




## Model terms

Another way to define the model is to define the arguments `model_terms` and defining the dependent variable with the argument `dep`. `model_terms` can be passed as a right-hand formula or as a list.

```{r warning=FALSE, message=FALSE, results='hide', eval=FALSE}
gamlj_lm(dep=ycont,model_terms=~x*z , data=data)
```

```{r warning=FALSE, message=FALSE, results='hide', eval=FALSE}
gamlj_lm(dep=ycont,model_terms=list("x","z",c("x","z")) , data=data)
```

When terms is passed as a list, each element is a term, and elements with more than 1 term, such as `c("x","z")` are interactions.

For mixed and generalized mixed models, this method requires to define the random effects of the model with argument `re`. 

For mixed and generalized mixed models, there are also the terms of the comparison model that can be defined as formula:

* nested_re: define the random coefficients of a nested model to compare.
* nested_terms: define the random coefficients of a nested model to compare.

```{r warning=FALSE, message=FALSE, results='hide', eval=FALSE}
gamlj_mixed(formula=ycont~x+(1+x|cluster),nested_terms=~1,nested_re=~1|cluster, data=data)
```


## List of variables

Finally, one can simply specify the factors and the covariates. In this way, however, no higher terms can be defined.

```{r warning=FALSE, message=FALSE, results='hide', eval=FALSE}
gamlj_lm(dep=ycont,factors=c("cat2"),covs=c("x","z") , data=data)
```

For mixed and generalized mixed models, this method requires to define the random effects of the model with argument `re`

# Passing other arguments

Several arguments if gamlj_* functions can be passed either as formula or as a list. They are:

* emmeans: estimate marginal means.
* posthoc: make post-hoc comparisons.
* nested_terms: define the terms of a nested model to compare.

For both argument, either a right-hand formula or a list can be passed.


```{r warning=FALSE, message=FALSE, results='hide', eval=FALSE}
data$cat2<-factor(data$cat2)
data$cat3<-factor(data$cat3)

gamlj_lm(formula=ycont~cat2*cat3,posthoc=~cat2+cat2:cat3 , data=data)
```

This produces the comparisons only across cat2 levels and the comparison across the interaction $cat2*cat3$. For these arguments, we do not use the formula `cat2*cat3` because it is not expanded, so passing `cat2*cat3` would produce two tables of comparisons, one for the comparison across cat2 levels, and one across cat3 levels. 

The equivalent argument as a list is:

```{r warning=FALSE, message=FALSE, results='hide', eval=FALSE}
gamlj_lm(formula=ycont~cat2*cat3,posthoc=list("cat2",c("cat2","cat3")) , data=data)
```






[back to top](#model-formula)
